Enhanced Material Point Method to face dynamic problems:\thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-26 main.tex
Version dia 27-3-2020

\maketitle
\begin{abstract}
  Material Point Method (MPM) has arise in the recent years as an
  alternative to Finite Element Method (FEM) under the large
  deformation regime. However, the simulation of shock waves
  propagation and other high frequency problems is still challenging
  under this approach due the incapability of the standard MPM time
  integration scheme to filter spurious noises. To overcome this
  limitation in this paper a explicit predictor-corrector integration
  method has been proposed. Its superior performance mitigates the
  presence of spurious oscillations with minimal dissipation in high
  frequency problems. Other source of numerical noise in the MPM is
  due to the lack of smoothness of the interpolation functions, and
  occurs due to material points crossing computational grid
  boundaries. This noise results in spurious local variations at the material
  points, where strain-stress fields are computed. This could invalidate
  the solution in one case, or damage it in another. To overcome it,
  this document adopts the \textit{local maximum-entropy approximation
    schemes} (LME) a robust substitute for the wide range of shape
  function in the MPM. Local \textit{max-ent} approximation may be regarded as a
  \textit{thermalization} of Delaunay triangulation which resolves the
  degenerate cases resulting from the lack or uniqueness of the
  triangulation. Furthermore, by modifying one user-defined parameter they
  are able to behave like finite elements or as a mesh-free
  method. This capability allows to face a wide range of physics
  with a single the shape function. Finally this paper demonstrates
  the performance of both improvements thorough numerical examples.    
  \keywords{LME \and MPM \and Explicit predictor-corrector \and Dynamic problems}
  % \PACS{PACS code1 \and PACS code2 \and more}
  % \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Since the proposal of the MPM by Sulsky {\it  et al.}
(1994)\cite{Sulsky1994} as a generalization to solids of the Fluid
Implicit Particle (FLIP) method \cite{Brackbill1986}. It popularity
has increased due to its ability to deal with large strain regime
without suffer mesh distortion inaccuracies.

However, this method suffers other kind of instabilities, such those
when material points crossing cell boundaries. This give rise to the
development of other interpolation techniques to overcome this
limitation such as the generalized interpolation material point method
(GIMP) Bardenhagen \& Kober (2004)\cite{Bardenhagen2004}, which has
demonstrate to have a good performance in the finite deformation
regime. However, in the absence of a regular grid, construction of the
weighting functions is only achieved at considerable effort and
computational cost. Furthermore, as it is based in a voxel
discretization, it is prone to suffer voxel domains overlap or gaps when the
material point mesh becomes irregular, which can introduce severe
inaccuracies. This is similar to the difficulty encountered by the
finite element methods due to element distortion.
A more robust alternative is the dual domain material point method
(DDMP) proposed by Zhang {\it et al.}
(2011)\cite{Zhang2011a}. Unfortunately this method shows an
unsatisfactory behaviour when particle/cell ratio
decreases~\cite{DHAKAL2016301}, therefore DDMP requires a large number
of particles needed for convergence, this
makes the method very expensive.

the B-Spline MPM
Tran {\it et al.} (2019)\cite{Tran2019a}

the Conservative Taylor Least Squares \textbf{CTLS} reconstruction Wobbes {\it et al.}
(2018)\cite{E_Wobbes_2018}.

Local \textit{max-ent} shape functions was first introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite elements (\textbf{FE}) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness or the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
mesh-free algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative,
satisfy the partition of unity property, and provide an exact
approximation for affine functions \cite{Arroyo2006}. This
approximation scheme has been proof to have a good performance under
the dynamic regime by other researchers like Navas {\it et al.}
(2018)\cite{Navas2018a} and Li {\it et al.} (2012)\cite{Li2012} for
Optimal Transportation Meshfree (\textbf{OTM}) method. And more
recently under the MPM framework by Wobbes {\it et
  al.}(2020)\cite{Wobbes2020} but limiting the $\gamma$ parameter to
zero.

This techniques are devoted to mitigate the
``grid crossing'' error. Nevertheless, in the presence of shock waves spurious
numerical noises appears despite of this using techniques
\cite{Tran2019e}.

These numerical inaccuracies also known as wiggles
are due to the time discretization technique. A simple approach to
face those spurious noises is to add a non-physical damping source to the
equilibrium equations. To avoid this, many researchers
has proposed alternative time integration schemes which reduce the
presence of high-frequency noises by filtering them. Some of these
attempts are the implicit GIMP (iGIMP) Charlton {\it et al.}
(2017)\nocite{Charlton2017} or the generalised-$\alpha$ scheme for the
MPM proposed by Tran \& Solowski (2019)\nocite{Tran2019e}.

On the other hand, the time integration scheme here proposed is an accommodation of
the Newmark a-form ($\gamma = 0.5$ and $\beta = 0$)\footnote{Notice
that these $\gamma$ and $\beta$ parameters does not have the same
meaning for the LME shape functions.} which is the central difference
explicit (CD). We have choose this method among other suitable
alternatives as Wilson {\it et al.} (1972)\nocite{Wilson1972} or Chung
\& Hulbert (1993)\nocite{Geranlized_alpha_1993} because it simplicity
and it good performance dealing with solids dynamics problems like those
presented in \cite{Navas2018a}.

The aim of this paper is to mitigate this spurious oscillations by the
employ of a suitable combination of the maximum-entropy (or local
\textit{max-ent}) $\gamma$ family shape functions, and the proposal of a explicit 
predictor-corrector scheme.

The article is organized as follows. Section \ref{sec:Notation}
briefly reviews the notation here employed. Next, Section \ref{sec:derivation-mpm}
is devoted to present briefly the governing equations of the elastic
problem, the variational formulation and the Galerkin procedure. In Section \ref{sec:epc-algor-mpm} an explicit predictor-corrector
time integration scheme for the MPM is proposed. Section
\ref{sec:local-max-ent} briefly reviews the local \textit{max-ent}
basis functions here employed. In Section \ref{sec:dynamic-benchmark}
a comparison between both methods MPM and OTM employing \textit{max-ent} shape
function is performed as well a parametric study for the $\gamma$
parameter. Both techniques will face to solve a classic dynamic
benchmark proposed by Dyka \& Ingel (1995)\cite{Dyka1995}. Finally,
conclusions and future research topics are exposed in Section \ref{sec:conclusions}.

\section{Brief note concerning the notation}
\label{sec:Notation}
In what follows, we will adopt the following convention. All the
physical variables involved in this problem are collected in Table
\ref{tab:notation-table}.

\begin{table}
  \centering
  \caption{Physical variables involved in the problem}
  \label{tab:notation-table}
  \begin{tabular}{c c c}
    $\rho$ & Density field & Scalar \\
    $\vect{a}$ & Acceleration field & First order tensor \\
    $\vect{v}$ & Velocity field & First order tensor \\
    $\vect{u}$ & Displacement field & Second order tensor \\
    $\tens{\sigma}$ & Cauchy stress tensor & Second order tensor \\
    $\tens{\varepsilon}$ & Cauchy strain tensor & Second order tensor \\
    $\tens{D}$ & Constitutive tensor & Fourth order tensor\\
  \end{tabular}
\end{table}

Three kind of subscript or superscript are used within paper. The
subscript $\qed_p$ is used to define a particle variable. While the
subscript $\qed_I$ is reserved in this notation for denoting nodal
variables. And finally, the superscript $\qed^{\psi}$ involves a virtual
magnitude. For the operators, the convention is : $\dot{\qed}$ and
$\ddot{\qed}$ for the first and second time derivative, $\qued \otimes
\qued$ means the dyadic operator, $(\cdot)$ and $(\colon)$ means the first
and second contraction of a tensor, $\Div{\qed}$
denotes the divergence operator, and finally $\Grad{\qed}$ and
$\GradS{\qed}$ denotes the gradient and its symmetric part. Einstain
subscripts convention is adopted so repeated index means addition.

\section{Derivation of the MPM}
\label{sec:derivation-mpm}
The aim of this section is to provide an overview of
the standard explicit MPM algorithm \cite{Sulsky1994}. Therefore, it
is structured in the following shape: we will first give a
introduction to the governing equations \ref{sec:governing-equations},
later the variational statement of the problem will be presented
\ref{sec:variational-formulation}, and finally a discretization
procedure thorough Galerkin is performed \ref{sec:Galerkin-procedure}.

The method has three main steps: (i) a variational recovery process,
where particle data is projected to the grid nodes, (ii) an Eulerian
step, where balance of momentum equation is expressed as a nodal
equilibrium equation thorough a FEM-like procedure, and finally (iii) a Lagrangian advection of the particles.

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. In a inifinitesimal
strain context, the behaviour of the continuum can be described by the
following set of equations:
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
(ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}},
\end{equation}
(iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
and finally (iv) the mass conservation, which can be obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div{\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual velocity field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{v}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{v}^{\psi} \in
  \mathcal{H}^1 \mid \vect{v}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{v}^{\psi}} < \infty\ \quad\text{and}\quad \Integral{3}{\tens{\varepsilon}^{\psi}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary value problem of elasticity is the function $\vect{v} \in
\mathcal{H}^1_0$ such that, for $\vect{v}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \Deriv{\vec{v}}{t}\ - \vec{b} \right) \cdot \vec{v}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{v}^{\psi}} - \Integral{3}{\tens{\sigma} \colon
    \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) is the balance of momentum formulated in
its weak form.

\subsection{Galerkin procedure}
\label{sec:Galerkin-procedure}
In order to arrive to a finite set of equations, in contrast with the
FEM, in the MPM a double discretization procedure is performed as we will
describe here below. First, the continuum domain
$\Omega$ is discretized with a finite sum of material points (in the
following particles), each one
represent a part of the discretized domain $\varOmega_p \subset
\varOmega$ with $p = 1,2\ldots ,N_p$ where $N_p$ is the number of
particles. The material point $\vec{x}_p$ is defined at the centroid
of each $\Omega_p$, figure \ref{fig:MPM-discretization}.

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \input{Figures/Discretization_MPM.tex}}
  \caption{MPM discretization.}
  \label{fig:MPM-discretization}
\end{figure}

% \begin{figure}\sidecaption
%   \centering
%   \resizebox{0.9\hsize}{!}{
%     \includegraphics[width=\textwidth]{./Figures/Discretization_MPM}
%   }
%   \caption{MPM discretization.}
%   \label{fig:MPM-discretization}
% \end{figure}

Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual field $\vect{v}^{\psi}_{p}$. So employing the definition of the
material integral, where we recover the Riemann integral definition as
an addition of a finite set of points, and their volumes are
interpreted as quadrature weights. Therefore individual terms in
\eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \vec{a} \cdot \vect{v}^{\psi}} =
    \dot{\vec{v}}_{p} \cdot \vect{v}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \colon \tens{\varepsilon}^{\psi}} =
   \tens{\sigma}_{p}\ \colon \tens{\varepsilon}^{\psi}_p \ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b} \cdot \vect{v}^{\psi} } = 
  \vec{b}_{p} \cdot \vect{v}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{v}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s \cdot \vect{v}^{\psi}} = \vec{t}^s_{p} \cdot \vect{v}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and divergences.

>>>>

\maketitle
\begin{abstract}
  Material Point Method (MPM) has arisen in the recent years as an
  alternative to Finite Element Method (FEM) under the large
  deformation regime. However, the simulation of shock waves
  propagation and other high frequency problems is still challenging
  under this approach due the incapability of the standard MPM time
  integration scheme to filter spurious noises. To overcome this
  limitation in this paper, an explicit predictor-corrector time integration
  scheme has been proposed. Its superior performance mitigates the
  presence of spurious oscillations with minimal dissipation in high
  frequency problems. Other source of numerical noise in the MPM 
  occurs when to material points cross computational grid
  boundaries and motivated due to the lack of smoothness of the interpolation
  functions. This noise results in spurious local variations at the material
  points, where strain-stress fields are computed. This could invalidate
  the solution in one case, or damage it in another. To overcome it,
  this document adopts the \textit{local maximum-entropy approximation
    schemes} (LME) a robust substitute for the wide range of shape
  function in the MPM. Local \textit{max-ent} approximation may be regarded as a
  \textit{thermalization} of Delaunay triangulation which resolves the
  degenerate cases resulting from the lack or uniqueness of the
  triangulation. Furthermore, by modifying a regularization parameter they
  are able to behave finite element like or as a mesh-free
  method. This capability allows to face a wide range of physics
  with a single shape function family. Finally this paper demonstrates
  the performance of both improvements thorough numerical examples.    
  \keywords{LME \and MPM \and Explicit predictor-corrector \and Dynamic problems}
  % \PACS{PACS code1 \and PACS code2 \and more}
  % \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Since the proposal of the MPM by Sulsky {\it  et al.}
(1994)~\cite{Sulsky1994} as a generalization to solids of the Fluid
Implicit Particle (FLIP) method~\cite{Brackbill1986}. It popularity
has increased due to its ability to deal with large strain regime
without suffer mesh distortion inaccuracies.

However, this method suffers other kind of instabilities, such those
when material points crossing cell boundaries. This give rise to the
development of other interpolation techniques to overcome this
limitation such as the generalized interpolation material point method
(GIMP) Bardenhagen \& Kober (2004)~\cite{Bardenhagen2004}, which has
demonstrate to have a good performance in the finite deformation
regime. However, in the absence of a regular grid, construction of the
weighting functions is only achieved at considerable effort and
computational cost.  Furthermore, as it is  a voxel based
discretization technique, it is prone to suffer voxel domains overlap
or gaps when the material point mesh becomes irregular, which can
introduce severe inaccuracies. This is similar to the difficulty
encountered by the finite element methods due to element distortion.
A more robust alternative is the dual domain material point method
(DDMP) proposed by Zhang {\it et al.}
(2011)~\cite{Zhang2011a}. Unfortunately this method shows an
unsatisfactory behaviour when particle/cell ratio
decreases~\cite{DHAKAL2016301}, therefore DDMP requires a large number
of particles needed for convergence, this
makes the method very expensive. In recent years the employ of
spline-lines has gain popularity with the introduction of the B-Spline
MPM proposed by Roel Tielen {\it et al.} (2017)~\cite{TIELEN2017265},
this technique allows the employ of unstructured set of notes and
particles. More recently, approximants derived from minimization has
been introduced in to the MPM framework with the Conservative Taylor
Least Squares (CTLS) reconstruction proposed by Wobbes {\it et al.}
(2018)~\cite{E_Wobbes_2018}, unfortunately when particles are spread in
a challenging way, the quality of the CTLS approximation decrease
locally.

This document adopts the \textit{local maximum-entropy approximation
    schemes} (LME) as a robust substitute for the wide range of shape
  function in the MPM. First introduced by Arroyo \& Ortiz
(2006)~\cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite element method (FEM) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness or the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
mesh-free algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative,
satisfy the partition of unity property, and provide an exact
approximation for affine functions~\cite{Arroyo2006}. This
approximation scheme has been proof to have a good performance under
the dynamic regime by other researchers like Navas {\it et al.}
(2018)~\cite{Navas2018a} and Li {\it et al.} (2012)~\cite{Li2012} for
Optimal Transportation Meshfree (OTM) method. And more
recently under the MPM framework by Wobbes {\it et
  al.}(2020)~\cite{Wobbes2020} but without exploring the benefits of
the regularization parameter $\beta$.

This techniques are devoted to mitigate the
``grid crossing'' error. Nevertheless, in the presence of shock waves spurious
numerical noises appears despite of this using
techniques~\cite{Tran2019e}. These numerical inaccuracies also known
as wiggles are due to inaccuracies in the time discretization technique.
A simple approach to face those spurious noises is to add a
nonphysical damping source to the equilibrium equations, this
approach has been widely employed in this and many other numerical
techniques. To avoid introducing this nonphysical sources, many
researchers has proposed alternative time integration
schemes which reduce the presence of high frequency noises by
filtering them or increasing the accuracy of the time integration
scheme. One of the most popular is the implicit GIMP (iGIMP)
Charlton {\it et al.} (2017)~\cite{Charlton2017}, more recently Tran \&
Solowski (2019)~\cite{Tran2019e} proposed a generalised-$\alpha$ scheme
for the MPM with promising results but at the expense of increasing
the computational effort. In this paper a less time consuming and high
efficient explicit predictor corrector integration method has been
proposed. It consists in an accommodation of the Newmark
predictor-corrector or central difference explicit (CD). We have
choose this method among other suitable alternatives as those proposed
by Wilson {\it et al.} (1972)~\cite{Wilson1972} or Chung \& Hulbert
(1993)~\cite{Geranlized_alpha_1993} because it simplicity and it good
performance dealing with solids dynamics problems under a mesh-free
framework in~\cite{Navas2018a}.

The aim of this document is to mitigate the spurious oscillations due
inaccuracies in both space and time discretization by the employ of a
suitable combination of the maximum-entropy (or local
\textit{max-ent}) family shape functions, and the proposal of
a explicit predictor-corrector scheme. We illustrate the advantages of
this approach thorough to simple but challenging test cases, like the
propagation of shock waves in a elastic bar or the response of a block of soil
gradually loaded with gravitational forces.

The article is organized as follows. Section \ref{sec:Notation}
briefly reviews the notation here employed. Next, Section \ref{sec:derivation-mpm}
is devoted to present briefly the governing equations of the elastic
problem, the variational formulation and the Galerkin procedure. In Section \ref{sec:epc-algor-mpm} an explicit predictor-corrector
time integration scheme for the MPM is proposed. Section
\ref{sec:local-max-ent} briefly reviews the local \textit{max-ent}
basis functions here employed. In Section
\ref{sec:Application-linear-elasticity-dynamic-problems} applications
to prove the numerical accuracy of the proposed approach are presented. Finally,
conclusions and future research topics are exposed in Section \ref{sec:conclusions}.

\section{Brief note concerning the notation}
\label{sec:Notation}
In what follows, we will adopt the following convention. All the
physical variables involved in this problem are collected in Table
\ref{tab:notation-table}. Three kind of subscript or superscript are used within paper. The
subscript $\qed_p$ is used to define a particle variable. While the
subscript $\qed_I$ is reserved in this notation for denoting nodal
variables. And finally, the superscript $\qed^{\psi}$ involves a
virtual magnitude. For the operators, the convention is : $\dot{\qed}$ and
$\ddot{\qed}$ for the first and second time derivative, $\qued \otimes
\qued$ means the dyadic operator, $(\cdot)$ and $(\colon)$ means the first
and second contraction of a tensor, $\Div{\qed}$ denotes the
divergence operator, and finally $\Grad{\qed}$ and $\GradS{\qed}$
denotes the gradient and its symmetric part. Einstain subscripts
convention is adopted therefore repeated index means addition.

\begin{table}
  \centering
  \caption{Physical variables involved in the problem}
  \label{tab:notation-table}
  \begin{tabular}{c c c}
    $\rho$ & Density field & Scalar \\
    $\vect{a}$ & Acceleration field & First order tensor \\
    $\vect{v}$ & Velocity field & First order tensor \\
    $\vect{u}$ & Displacement field & First order tensor \\
    $\vect{x}$ & Global coordinates & First order tensor \\
    $\vect{\xi}$ & Local coordinates & First order tensor \\
    $\tens{\sigma}$ & Cauchy stress tensor & Second order tensor \\
    $\tens{\varepsilon}$ & Cauchy strain tensor & Second order tensor \\
    $\tens{D}$ & Constitutive tensor & Fourth order tensor\\
  \end{tabular}
\end{table}

\section{Derivation of the MPM}
\label{sec:derivation-mpm}
The aim of this section is to provide an overview of the standard
explicit MPM algorithm~\cite{Sulsky1994}. In consequence it
is structured as follows: the governing equations will be introduced
in \ref{sec:The boundary-value-problem}, later the variational statement of
the problem will be presented \ref{sec:variational-formulation}, and
finally a discretization procedure thorough Galerkin is performed in
\ref{sec:Galerkin-procedure}. The method has three main steps: (i) a
variational recovery process, where particle data is projected to the
grid nodes, (ii) an Eulerian step, where balance of momentum equation
is expressed as a nodal equilibrium equation thorough a FEM-like
procedure, and finally (iii) a Lagrangian advection of the
particles. In consequence the MPM can be regarded as a
Lagrangian-Eulerian method where particles carries on all the physical
information and a set of background nodes is employed to
compute the equilibrium equation.

\subsection{The boundary-value problem}
\label{sec:The boundary-value-problem}
In MPM the continuum mechanics approach is considered. So on, let
define a continuum domain $\varOmega$ occupied by an elastic body like the sketched in
\ref{fig:Continuum-solid}, and $\partial \varOmega$ the boundaries of
the domain defined by $\partial \varOmega = \Gamma_d \bigcup
\Gamma_n$ and $\Gamma_d \cap  \Gamma_n = \emptyset$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}\sidecaption
  \centering
  \resizebox{0.8\hsize}{!}{
    \includegraphics[width=\textwidth]{Figures/Continuum-solid}}
  \caption{Description of the boundary-value-problem in a
    continuum. Red lines represents the closure $\partial \varOmega$
    of the domain $\varOmega$ represented in gray.}
  \label{fig:Continuum-solid}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this context the field $\vect{u}$ allows to describe the \textit{global state}
of the system. Now the variable $\phi =
(\tens{\varepsilon},\tens{\sigma})$ is defined as the set of \textit{local
  states} at any point of the continuum which can be derived from the
field $\vect{u}$ through the following set of governing equations and
restrictions that must be satisfied. First (i) the \textit{compatibility
  equation} that extracts from $\vect{u}$ the strain field is,
\begin{equation}
  \label{eq:Compatibility-equation}
  \tens{\varepsilon} = \GradS{\vect{u}},
\end{equation}
together with essential boundary conditions of Dirichlet type
$\varGamma_d$. An additional consideration over the strain field is
the assumption of inifinitesimal strain, therefore second order terms
in the spatial derivatives can be neglected. The corresponding
conjugate variable for the strain field if the stress field
$\tens{\sigma}$, which satisfies (ii) the \textit{conservation of
  momentum equation}
\begin{equation}
  \label{eq:Balance-momentum}
\rho \frac{D\vect{v}}{Dt} = \Div{\tens{\sigma}} + \rho \vect{b}
\end{equation}
together with the natural boundary conditions of the Neumann type
$\varGamma_n$. Next (iii) the constitutive equation as a linear
application from $\Re^n$ to  $\Re^n$, which relates the
strain tensor with the stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\tens{\sigma} = \tens{D} \colon \tens{\varepsilon}.
\end{equation}
The final restriction is (iv) the mass conservation, which can be obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div{\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual displacement field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{u}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{u}^{\psi} \in
  \mathcal{H}^1 \mid \vect{u}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{u}^{\psi}} < \infty\ \quad\text{and}\quad
  \Integral{3}{\tens{\varepsilon}^{\psi}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary-value problem of elasticity is the function $\vect{u} \in
\mathcal{H}^1_0$ such that, for $\vect{u}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \frac{d\vec{v}}{dt}\ - \vec{b} \right) \cdot \vec{u}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{u}^{\psi}} - \Integral{3}{\tens{\sigma} \colon
   \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) together with
\eqref{eq:Constitutive-equation} and
\eqref{eq:Rho-material-derivative} represents the weak form
formulation of the problem.

\subsection{Galerkin procedure}
\label{sec:Galerkin-procedure}
In order to arrive to a finite set of equations, in contrast with the
FEM, in the MPM a double discretization procedure is performed as we will
describe here below. First, the continuum domain
$\Omega$ is discretized with a finite sum of material points (in the
following particles), each one
represent a part of the discretized domain $\varOmega_p \subset
\varOmega$ with $p = 1,2\ldots ,N_p$ where $N_p$ is the number of
particles. The material point $\vec{x}_p$ is defined at the centroid
of each $\Omega_p$, figure \ref{fig:MPM-discretization}.
Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual displacement field $\vect{u}^{\psi}_{p}$. Therefore, employing
the definition of the material integral, where we recover the Riemann
integral definition as an addition of a finite set of points, and
their volumes are interpreted as quadrature weights. Consequently,
individual terms in \eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \frac{d\vec{v}}{dt} \cdot \vect{u}^{\psi}} =
    \frac{d\vec{v}_{p}}{dt} \cdot \vect{u}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \colon \tens{\varepsilon}^{\psi}} =
   \tens{\sigma}_{p}\ \colon \tens{\varepsilon}^{\psi}_p \ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b} \cdot \vect{u}^{\psi} } = 
  \vec{b}_{p} \cdot \vect{u}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{u}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s \cdot \vect{u}^{\psi}} = \vec{t}^s_{p} \cdot \vect{u}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and
divergences.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}\sidecaption
  \centering
  \resizebox{0.8\hsize}{!}{
    \includegraphics[width=\textwidth]{Figures/Mesh-particles-back}}
  \caption{Description of the spatial discretization for domain presented in the
    figure \ref{fig:Continuum-solid}. Blue mesh represent the
    background computational support, and the grid mesh conforms the
    discretized continuum body.}
  \label{fig:MPM-discretization}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Introducing \eqref{eq:particle_acceleration_forces},
\eqref{eq:particle_internal_forces}, \eqref{eq:particle_body_forces}, 
and \eqref{eq:particle_load_forces} in \eqref{eq:BalanceMomentum_wf},
and approximating the displacement field of the particle
$p$ as $\vec{u}_{p} = N_{Ip} \vect{u}_I$, $\vec{u}^{\psi}_{p} = N_{Ip} \vect{u}_I^{\psi}$,
and its gradient as $\tens{\varepsilon}_p = (\vect{u}_I \otimes
\Grad{N_{Ip}})^s$, $\tens{\varepsilon}_p^{\psi}= (\vect{u}_I^{\psi} \otimes \Grad{N_{Ip}})^s$.
We reach to the nodal balance of forces of the continuum,
\begin{equation}
  \label{eq:particle_balance_forces3}
  \dot{\vec{p}}_{I}= \tens{m}_{IJ}\dot{\vec{v}}_{J} = \vec{f}_{I}^{int} + \vec{f}_{I}^{ext},
\end{equation}
where $\dot{\vec{p}}_{I}$ is the rate of momentum at grid node $I$, the nodal mass matrix $\tens{m}_{IJ}$ is,
\begin{equation}
  \label{eq:particle_nod_mass}
  \tens{m}_{IJ} = N_{Ip} m_p N_{Jp}.
\end{equation}
To improve the computational efficiency and stability, the nodal mass matrix
\eqref{eq:particle_nod_mass} can be substituted by the lumped mass
matrix $\tens{m}_{IJ}^{lumped}$.
Later, internal and external forces are computed as follows,
\begin{equation}
  \label{eq:nodal_internal_forces}
  \vec{f}_{I}^{int} = - \tens{\sigma}_{p} \cdot \Grad{N_{Ip}} \frac{m_p}{\rho_p}
\end{equation}
\begin{equation}
  \label{eq:nodal_external_forces}
  \vec{f}_{I}^{ext} = N_{Ip}\ \vec{b}_{p}\ m_p  + N_{Ip}\ \vec{t}^s_{p}\ m_p h^{-1} 
\end{equation}
where $\tens{\sigma}_{p} = \tens{\sigma}_{p}(\tens{\varepsilon}_{p})$
is the particle $p$ stress field, which can be integrated employing
the suitable constitutive model. The strain tensor is updated employing the rate of stress tensor $\dot{ \tens{\varepsilon}}_{p}$ used to update the
strain tensor is as follows \eqref{eq:IncrStrainPoint}.
\begin{equation}
  \label{eq:IncrStrainPoint}
  \dot{\tens{\varepsilon}_{p}} = \frac{\Delta
    \tens{\varepsilon}_{p}}{\Delta t} =
  \frac{1}{2} \left[\Grad{N_{Ip}}\ \otimes \vec{v}_{I} + \vec{v}_{I} \otimes
    \Grad{N_{Ip}}\ \right].
\end{equation}
